---
title: "Bayesian Metanalysis"
output: html_notebook
---


```{r setup}
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(skimr)
library(readxl)
library(here)
library(tidyverse)
library(tidybayes)
library(bayesplot)
library(brms)
```

```{r}
convert_czech_numeric <- function(column) {
  result <- as.numeric(gsub(",",".", column, fixed = TRUE))
  if(any(is.na(result) != is.na(column))) {
    stop("Problem converting")
  }
  result
}

data <- read_excel(here("private_data","--META-DATA-5.xlsx"), sheet = "DATA 5") %>%
  rename(case_no = "source case n.", mut_nucl = "mu nucl", mut_amino = "principal mutation", mut_type = "mu-type", additional_mutation = "additional mutation", REPROD = "Reprod. organs abnorm") %>%
  select(source, case_no,gene, mut_nucl, mut_amino, mut_type, 
         Sex, Age, RD:speech) %>%
  mutate(CI = convert_czech_numeric(CI), DD = convert_czech_numeric(DD),
         LIV = convert_czech_numeric(LIV), REN = convert_czech_numeric(REN),
         REPROD = convert_czech_numeric(REPROD),
         Sex = factor(toupper(Sex)), 
         source = factor(source),
         gene = factor(gene))

data %>% skim()
```

```{r}
data %>% group_by(gene) %>% summarise(count = length(gene))
```


```{r}
phenotypes_to_use <- c("RD","OBE","PD","CI","REN")
genes_to_use <- paste0("BBS",1:8)

data_long <- data %>% 
  rowid_to_column("ID") %>%  
  gather("phenotype","phenotype_value",RD:speech) %>%
  filter(phenotype %in% phenotypes_to_use, gene %in% genes_to_use) %>%
  filter(!is.na(phenotype_value)) %>%
  mutate(phenotype_value = as.integer(if_else(phenotype_value == 0, 0 , 1)))

data_long %>% group_by(gene, phenotype) %>% summarise(count = length(phenotype_value), prop = mean(phenotype_value))
```

TODO co s pulkami?

Formula 1 + (0+gene)|phenotype - korelace je 8x8, tj. vliv na renalni dispatii je korelovany mezi geny (což nechceme)

```{r}
formula_gene_only <- brmsformula(phenotype_value ~ 1 + (0 + phenotype)|gene , family = "bernoulli")

#get_prior(formula_gene_only, data = data_long)
# priors <- c(prior(normal(0,2), class = Intercept),
#             prior(normal(0,2), class = sd)
#             )

#make_stancode(formula_gene_only, data_long, prior = priors)

fit_gene_only <- brm(formula_gene_only, prior = priors, data = data_long) #, control = list(adapt_delta = 0.9)
fit_gene_only
```



```{r}
run_pp_checks <- function(fit, data_long, out_func = print) {
  pp_check(fit, "bars", nsamples = 500, prob = 0.95, fatten = 1.5) %>% out_func
  pp_check(fit, "bars_grouped", nsamples = 500, group = "gene", prob = 0.95, fatten = 1.5)%>% out_func
  pp_check(fit, "bars_grouped", nsamples = 500, group = "phenotype", prob = 0.95, fatten = 1.5)%>% out_func
  
  predicted <- posterior_predict(fit) 
  ppc_bars_grouped(data_long$phenotype_value, predicted, group = if_else(is.na(data_long$Sex), "NA", as.character(data_long$Sex)), prob = 0.95, fatten = 1.5) %>% out_func
}
```

```{r}
run_pp_checks(fit_gene_only, data_long)
```



```{r}


```

```{r}

```


```{r}
launch_shinystan(fit_gene_only)
```


The model:
Individual phenotypes are predicted by:
  * The mutated subdomain (hierarchical)
    * Possibly a hierarchical effect of the specific mutation
  * Sex
  * Study (hierarchical)

TODO: Figure out a way to make phenotypes correlated - maybe have hierarchical MVN priors for the coeffs?