---
title: "Additional multiverse analysis"
output: html_notebook
---

TODO: 
- Rozdily v RD zmizi pri imputation Sex
- min_probable_OR ma divný scale
- podivat se na "divne" geny (případně vyhodit)
- lof -> cLOF

Text:
- Upravit text o filtraci (vyrazene fenotypy) - nebo spustit jeste nekdy se vsemi feno...


This describes all Bayesian models we tried throughout this project. 

Further we discuss the reasoning behind our choice of model for the main analysis, in particular why between-study variability is crucial while age, sex and loss of function can be omitted.


Finally we compute how the conclusions of the paper hold under all of the models.

# Model description

All models are logistic regression using `brms`.

```{r setup}
library(rstan)
library(brms)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(skimr)
library(readxl)
library(here)
library(mice)
library(tidyverse)
library(tidybayes)
library(bayesplot)
library(cowplot)

source(here("data_processing.R"))
source(here("models.R"))
source(here("models_funcs.R"))
source(here("plots.R"))

##PP Checks take time, so turn this off for quicker computation of the rest
compute_pp_checks <- FALSE
```


```{r load_data}
data <- read_main_data()
genes_to_show <- genes_to_show_from_data(data)
data_long <- data_long_from_data(data)
```

## Base models

Base models are those that work with (a subset of) the original dataset, without any imputation. The models are defined in file `models.R`. They differ in the model formula, subsets of the dataset they use and priors for model coefficients. Here is the list of base models.

```{r}
models_base %>% walk(function(def) {
  cat(def$name,":\n\tformula:", as.character(def$formula)[1], "\n\tdata filter: ", def$filter)
  if(def$note != "") {
    cat("\tNote:", def$note)
  }
  cat("\n\n")
  #tibble(name = def$name, formula = as.character(def$formula)[1], filter = def$filter,  note = def$note)
})
```


```{r fit_base_models}
fits_base <- models_base %>%
  map(function(def) {
     fit_base_model(def, data_long)
  }
) 
```


```{r}
# data_for_prediction_gene_only <- tibble(gene = genes_to_show) %>%
#     crossing(tibble(phenotype = phenotypes_to_show)) %>% mutate(functional_group = functional_group_for_gene(gene))
# 
# data_for_prediction_gene_only_BBSome <- data_for_prediction_gene_only %>% filter(functional_group_for_gene(gene) == "BBSome", gene != "BBS18")
# 
# 
# data_for_prediction_gene_source <- tibble(gene = genes_to_show, source = "new_source") %>%
#     crossing(tibble(phenotype = phenotypes_to_show)) %>% mutate(functional_group = functional_group_for_gene(gene))
# 
# data_for_prediction_gene_source_BBSome <- data_for_prediction_gene_source %>% filter(functional_group_for_gene(gene) == "BBSome", gene != "BBS18")
# 
# data_long_filter_lof <- data_long %>% filter(loss_of_function == "certain")
# 
# data_for_prediction_gene_source_lof <- tibble(gene = genes_to_show, source = "new_source", loss_of_function_certain = 1) %>%
#     crossing(tibble(phenotype = phenotypes_to_show)) %>% mutate(functional_group = functional_group_for_gene(gene))
# 
# plot_pairwise_differences(fit_gene_source_age_imputed, data_for_prediction_gene_source_BBSome %>% crossing(age_std_for_model = age_transform(c(10,30,50))))
# 
# plot_pairwise_differences(fit_gene_source_age_sex_imputed, data_for_prediction_gene_source_BBSome %>% crossing(tibble(Sex = c("M","F"))) %>% crossing(age_std_for_model = age_transform(c(10,30,50))))

```

```{r}
# fit <- fit_base_model(models_base$gene_source_filtered_sex, data_long)
# data_for_prediction <- data_for_prediction_base_model(models_base$gene_source_filtered_age, genes_to_show, phenotypes_to_show)
# 
# plot_gene_phenotype_differences_estimates(fit, data_for_prediction)
```




## Imputation with mice

Including age or sex in the base models is problematic as it involves tossing out `r round(100 * mean(is.na(data$age)))`% or `r round(100 * mean(is.na(data$Sex)))`% of data respectively. This results in wide posterior intervals and weak inferences. To try to ameliorate this we also tested running models on datasets with age and sex imputed, using multiple imputation via the `mice` package. We assume that both age and sex can be realted to the functional group of the mutation and to each other. Involving further relations (e.g. individual genes) led to warnings from the `mice` package and we thus didn't use those.

```{r mice_impute}
set.seed(20181217)
predictor_matrix <- matrix(0, nrow = 2, ncol = ncol(data))
colnames(predictor_matrix) <- names(data)

predictor_matrix[, c("functional_group")] <- 1
predictor_matrix[1, c("Sex")] <- 1
predictor_matrix[2, c("age_std_for_model")] <- 1

mice_m = 5
data_mice <- mice(data, m = mice_m, blocks = list("age_std_for_model", "Sex"), predictorMatrix = predictor_matrix)
if(!is.null(data_mice$loggedEvents)) {
  data_mice$loggedEvents %>% select(dep, meth, out) %>% distinct()
}

data_long_mice <- mice::complete(data_mice, "all") %>% map(data_long_from_data)
```

The models differ in the formulas used, but all use the default priors and do not filter the dataset in any way.

## Fit the imputed models

```{r fit_imputed_models}
fits_imputed <- models_imputed %>% map(function(def) { fit_imputed_model(def, data_long_mice)})
all_fits <- c(fits_base, fits_imputed)
```


# Between-study variability needs to be included

# Between-study variability mostly explains age and sex differences

# Between-study variability mostly explains loss-of-function differences

## PP Check for LOF

```{r}
compare_pp_check <- function(model_names, pp_check_types) {
  for(name in model_names) {
    run_pp_checks(all_models[[name]], all_fits[[name]], data_long, types = pp_check_types)  
  }
} 
```


```{r, fig.width=8}
models_to_compare = c("gene_source", "gene_source_lof", "gene_source_filtered_lof", "gene_lof")

if(compute_pp_checks) {
  compare_pp_check(models_to_compare, c("gene_lof"))
}
```

First, let us note that the checks for `gene_source` and `gene_source_lof` are almost identical. So let us focus on the comparison between `gene_source` and `gene_source_filtered_lof`.

What we see is that the `gene_source` model has trouble explaining some properties of unknown.BBS01, BBS02, unknown.BBS03, certain.BBS06 (just barely), certain.BBS07, certain.BBS08 (just barely), certain.BBS10, unknown.BBS12, certain.other.

`gene_source_filtered_lof` only slightly improves over this some of the problematic genes became barely OK and others stayed the same (obviously it does not model the uncertain LOF data).

`gene_lof` improves in certrain.BBS08 and certain.other while having problems with unknown.BBS09 (barely) and with slightly worse fit for unnknown.BBS03, unknown.BBS07.

So while there improvements in fitting LOF differences per gene when including LOF in the model, they are not very strong.

```{r}
if(compute_pp_checks) {
  compare_pp_check(models_to_compare, c("phenotype_lof"))
}
```

Once again `gene_source` and `gene_source_lof` are almost identical. Overall the phenotype comparison is worse for the `gene_source` model as it has problems with OBE (barely), certain.CI (barely), REN, LIV, while `gene_source_filtered_lof` has no problems. `gene_lof` is similar to `gene_source` but improves for LIV.

The substantial problems with fitting REN in `gene_source` are potentially troubling.

This nevertheless indicates that there is some missing information in all fits. 

```{r, fig.width=8}
if(compute_pp_checks) {
  compare_pp_check(models_to_compare, c("source"))
}
```

`gene_source` and `gene_source_lof` are once again almost identical.

The comparison between `gene_source` and `gene_lof` is not clear as for sources there is improvement in one and for some in the other.

`gene_source_filtered_lof` removes some of the issues in `gene_source` (e.g. Scheidecker et al. 2013), but retains others (Chaki et al., Schaefer et al. 2011 (a), Schaefer et al. 2016.).

Once again there is a minor advantage in filtering out LOF data while there is no advantage to including LOF in the model. There are at least two possible explanations: when filtering for LOF, there is much less data, but almost the same number of parameters, so the model can accommodate (or even slightly overfit) the data. Another plausible explanation is that there is some important variability among the patients with mutations that are not certainly LOF which hinder fitting.  We however couldn't devise a defensible way to further split the mutations that are not certainly LOF. 

```{r, fig.width=10, fig.height=7}
plot_pairwise_differences(models_base$gene_source_filtered_lof, fits_base$gene_source_filtered_lof, data_for_prediction_base_model(models_base$gene_source_filtered_lof, genes_to_show, phenotypes_to_show) %>% filter_for_BBSome())
```


TODO

In most cases, the loss of function does not make for a big difference in predictions. This is because there is no strong distinction in the data itself:

```{r, fig.width=8, fig.height=4}
lof_data_plot <- data_long %>% 
  group_by(source, phenotype, gene, loss_of_function) %>% 
  filter(functional_group == "BBSome", gene != "BBS18") %>%
  summarise(n_patients = length(phenotype_value), proportion = sum(phenotype_value) / n_patients) %>%
  ggplot(aes(x = gene, y = proportion, size = n_patients, color = loss_of_function)) +
  geom_jitter(alpha = 0.3, height = 0, width = 0.2) +
  facet_wrap(~phenotype, nrow = 2) +
  scale_y_continuous("Proportion of phenotype") +
  scale_size_continuous(range = c(0.5,4)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust =0.5))

lof_data_plot
```

# Some models are a bad fit for the data

gene_only
gene_only_filtered_lof
gene_lof
gene_lof_per_gene
gene_source_lof ??
very narrow

# Robustness of conclusions to model choice

Definovat min. klin. relevantni efekt

```{r}
min_interesting_effect <- 2
min_phenotypes_for_diffs <- 3
min_bbsome_pairwise_for_diffs <- 5

evaluators <- list()
```




```{r}
n_genes_lof_different <- 5
```

*	The severity of BBS is worse in patients with LOF mutations than in patients with other mutations. 
  * Ignored for models that do not include LOF
  * Measured as posterior probability, that the LOF effect is positive (OR > 1) for at least `r n_genes_lof_different` BBSome genes in `r min_phenotypes_for_diffs` phenotypes.
```{r}
#TODO: look on predictions for LOF

evaluator_lof <- function(def, fit) {
 if("lof" %in% def$additional_components || "lof per gene" %in% def$additional_components)  {
   # phenotype_coeffs <- paste0("phenotype", phenotypes_to_use,":loss_of_function_certain")
   # genes_to_check <- sprintf("BBS%02d",c(1,2,4,5,7,8,9))
   # samples_raw <- coef(fit, summary = FALSE)$gene[,genes_to_check,phenotype_coeffs]
   # #grid <- list(sample = 1:(dim(samples_raw)[1]), gene = genes_to_check, phenotype = phenotypes_to_use)
   # grid <- list(sample = 1:(dim(samples_raw)[1]), gene_id = 1:length(genes_to_check), phenotype_id = 1:length(phenotypes_to_use))
   # samples_tidy <- expand.grid(grid)
   #samples_tidy$lof_coef <- samples_raw[as.matrix(samples_tidy)]
   #  samples_tidy %>% mutate(gene = genes_to_check[gene_id], phenotype = phenotypes_to_use[phenotype_id])
   
   genes_to_check <- sprintf("BBS%02d",c(1,2,4,5,7,8,9))
   data_for_prediction_lof <- data_for_prediction_base_model(def, genes_to_check, phenotypes_to_show = phenotypes_to_show, age_transform = age_transform_from_age(data$age_numbers_groups_guessed))
   samples_tidy <- get_samples_lof_diff(fit, data_for_prediction_lof)
   
   samples_tidy %>% 
     group_by(sample, phenotype) %>% summarise(enough_genes = sum(odds_ratio > 1) >= n_genes_lof_different) %>%
     group_by(sample) %>% summarise(enough_phenotypes = sum(enough_genes) >= min_phenotypes_for_diffs) %>%
     ungroup() %>% summarise(`cLOF mutations more severe` = mean(enough_phenotypes))
 } else {
   tibble(`cLOF mutations more severe` = NA_real_)
 }
}
```
  
*	The data suggest a difference between the severity of BBS in patients with mutations in different functional groups of genes.
  * Measured as posterior probability, that there is clinically relevant effect for at least `r min_phenotypes_for_diffs` phenotypes for at least 1/2 of pairwise gene comparisons.
  
```{r}
evaluator_functional_groups_differ <- function(group1, group2) {
  function(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
    target_column_name = paste(bbs_labeller(group1),"and",bbs_labeller(group2),"differ")
    samples_diff %>% 
      filter(functional_group.x == group1, functional_group.y == group2) %>%
      group_by(sample, phenotype) %>%
      summarise(enough_pairwise_relevant = mean(is_relevant) >= 1/2) %>%
      group_by(sample) %>%
      summarise(enough_phenotype_relevant = sum(enough_pairwise_relevant) >= min_phenotypes_for_diffs) %>%
      ungroup() %>%
      summarise(!!target_column_name := mean(enough_phenotype_relevant))  
  }
}

evaluators[["bbs3_vs_bbsome"]] <- evaluator_functional_groups_differ("BBS03","BBSome")
evaluators[["bbs3_vs_chaperonins"]] <- evaluator_functional_groups_differ("BBS03","Chaperonins")
```
*	The data suggest a difference between the severity of BBS in patients with mutations in different BBSome subunits.
  * Measured as posterior probability, that there is clinically relevant effect for at least `r min_phenotypes_for_diffs` phenotypes for at least `r min_bbsome_pairwise_for_diffs` pairwise comparisons. 
```{r}
evaluators[["bbsome_genes_differ"]] <- function(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
samples_diff_bbsome %>% 
  filter(as.character(gene.x) > as.character(gene.y)) %>%
  group_by(sample, phenotype) %>%
  summarise(enough_pairwise_relevant = sum(is_relevant) >= min_bbsome_pairwise_for_diffs) %>%
  group_by(sample) %>%
  summarise(enough_phenotype_relevant = sum(enough_pairwise_relevant) >= min_phenotypes_for_diffs)  %>%
  ungroup() %>%
  summarise(`BBSome genes differ` = mean(enough_phenotype_relevant))
}
```

```{r}
    # def <- models_base$gene_source
    # fit <- all_fits[[def$name]]
    # data_for_prediction <- data_for_prediction_base_model(def, genes_to_show, phenotypes_to_show, age_transform = age_transform_from_age(data$age_numbers_groups_guessed))
    # 
    # data_for_prediction_bbsome <- data_for_prediction %>% filter_for_BBSome()
    # 
    # samples <- get_tidy_samples(fit, data_for_prediction)
    # samples_diff <- get_samples_pairwise_diff(def, samples) %>% mutate(is_relevant = odds_ratio >= min_interesting_effect | odds_ratio <= 1 /min_interesting_effect)
    # 
    # samples_diff_bbsome <- get_tidy_samples(fit, data_for_prediction_bbsome) %>% get_samples_pairwise_diff(def, .) %>% mutate(is_relevant = odds_ratio >= min_interesting_effect | odds_ratio <= 1 /min_interesting_effect)
    # 
    # samples_prediction_bbsome = get_tidy_samples_prediction(fit, data_for_prediction_bbsome)
    # 
    # samples_linear_bbsome <- get_tidy_samples(fit, data_for_prediction_bbsome, scale = "linear")

```


*	BBS2 and BBS7 share similar pattern of penetrance of symptoms.
*	BBS4 phenotype is the most severe of all BBSome subunits.
  * Očekávám, že nebude potvrzeno (kvůli RD, OBE, PD)
  * Probability that BBS4 has top 3 odds ratios for at least 5 phenotypes
  * Probability that a patient with BBS4 has the most #phenotypes.
```{r}
evaluators[["bbs04_top"]] <- function(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
samples_diff_bbsome %>% 
    #TODO phenotype
  filter(gene.x == "BBS04", gene.y != "BBS04") %>%
  group_by(sample, phenotype) %>%
  summarise(bbs04_top = sum(odds_ratio < 1) <= 2) %>%
  group_by(sample) %>%
  summarise(bbs04_top_all = sum(bbs04_top) >= 5 ) %>%
  summarise(`BBS4 top odds` = mean(bbs04_top_all))
}
```
```{r}
evaluators[["bbs04_most_pheno"]] <- function(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
  samples_prediction_bbsome %>% 
    group_by(sample, gene) %>%
    summarise(num_phenotypes = sum(value)) %>%
    group_by(sample) %>%
    mutate(has_most_pheno = num_phenotypes >= max(num_phenotypes) ) %>%
    ungroup() %>%
    filter(gene == "BBS04") %>%
    #group_by(gene) %>%
    summarise(`BBS4 most #phenotypes` = mean(has_most_pheno))
}
```
*	Mutations in different BBSome subunits predispose to different renal phenotype.
  *  Measured as posterior probability, that there is clinically relevant effect for at least `r min_bbsome_pairwise_for_diffs` pairwise comparisons. 
```{r}
evaluators[["REN_bbsome_genes_differ"]] <- function(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
  samples_diff_bbsome %>% 
    filter(phenotype == "REN", as.character(gene.x) > as.character(gene.y)) %>%
    group_by(sample) %>%
    summarise(enough_pairwise_relevant = sum(is_relevant) >= min_bbsome_pairwise_for_diffs) %>%
    ungroup() %>%
    summarise(`REN: BBSome genes differ` = mean(enough_pairwise_relevant))
}
```

  
*	Polydactyly is more frequent in BBS2 patients compared to BBS1 patients.
  * Directly the probability that the OR > 1 (nebo OR > min. efekt)
```{r}
evaluator_prob_pairwise_diff <- function(phenotype_name, gene_more_frequent, gene_less_frequent) {
  function (samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {
     target_column_name = paste0(phenotype_name,": ", bbs_labeller(gene_more_frequent), " more frequent than ", bbs_labeller(gene_less_frequent))
      samples_diff %>% 
        filter(phenotype == phenotype_name, gene.x == gene_more_frequent, gene.y == gene_less_frequent) %>%
        group_by(sample) %>%
        summarise(all_higher = all(odds_ratio >= 1)) %>%
        ungroup() %>%
        summarise(!!target_column_name := mean(all_higher))   
  }
}

evaluators[["PD_BBS02_BBS01"]] <- evaluator_prob_pairwise_diff("PD","BBS02","BBS01")
```

* Polydactyly is more frequent in BBS10 patients compared to BBS1 patients
```{r}
evaluators[["PD_BBS10_BBS01"]] <- evaluator_prob_pairwise_diff("PD","BBS10","BBS01")

```

*	Cognitive impairment is less frequent in BBS3 patients compared to other patients (all canonical BBS genes).
  * Tohle trochu dubluje tu diskuzi o func. groups
  * Probability that all pairwise odds are less
```{r}
evaluator_prob_group_diff_all <- function(phenotype_name, group1, group2, group1_name = paste0(group1, collapse = "_"),group2_name = paste0(group2, collapse = "_")) {
    function (samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome) {

      target_column_name = paste0(phenotype_name,": ",bbs_labeller(group1_name)," more frequent than ", bbs_labeller(group2_name))
      samples_diff %>% 
        filter(phenotype == phenotype_name, gene.x %in% group1, gene.y %in% group2) %>%
        group_by(sample) %>%
        summarise(all_higher = all(odds_ratio >= 1)) %>%
        ungroup() %>%
        summarise(!!target_column_name := mean(all_higher))   
    }
}

evaluators[["CI_BBSO3_all"]] <- evaluator_prob_group_diff_all("CI", paste0("BBS0", c(1,2,4,5,7,8,9)), "BBS03", group1_name = "BBSome")
```

*	Cognitive impairment is more frequent in BBS7 patients compared to other patients wirh mutations in BBSome subunits.
  * Očekávám, že nebude potvrzeno
```{r}
evaluators[["CI_BBS07_all"]] <-  evaluator_prob_group_diff_all("CI", "BBS07", paste0("BBS0", c(1,2,4,5,8,9)), group2_name = "others")

```

*	Renal involvement is less frequent in BBS1, BBS4 and BBS8 patients compared to BBS2, BBS7, BBS9 patients.
```{r}
evaluators[["REN_groups_all"]] <- evaluator_prob_group_diff_all("REN", paste0("BBS0", c(2,7,9)),  paste0("BBS0", c(1,4,8)), group1_name = "BBS2,7,9", group2_name = "BBS1,4,8")
```

*	Renal involvment is less frequent in BBS1 patients compared to BBS2 patients.
```{r}
evaluators[["REN_BBS1_BBS2"]] <- evaluator_prob_pairwise_diff("REN","BBS02","BBS01")
```
*	Renal involvment is less frequent in BBS1 patients compared to BBS10 patients.
```{r}
evaluators[["REN_BBS1_BBS10"]] <- evaluator_prob_pairwise_diff("REN","BBS10","BBS01")
```
*	Liver involvment is less frequent in BBS1 patients compared to BBS2 patients.
```{r}
evaluators[["LIV_BBS1_BBS2"]] <- evaluator_prob_pairwise_diff("LIV","BBS02","BBS01")
```
*	Liver involvment is less frequent in BBS1 patients compared to BBS10 patients.
```{r}
evaluators[["LIV_BBS1_BBS10"]] <- evaluator_prob_pairwise_diff("LIV","BBS10","BBS01")
```
*	Patients with BBS2 mutations are more likely to have heart anomalies compared to patients with other BBSome mutations.
```{r}
evaluators[["HEART_BBS2_all"]] <-  evaluator_prob_group_diff_all("HEART", "BBS02", paste0("BBS0", c(1,4,5,7,8,9)), group2_name = "others")
```
*	Patients with BBS5 mutations are more likely to have liver anomalies compared to patients with other BBSome mutations.
```{r}
evaluators[["LIVER_BBS5_all"]] <-  evaluator_prob_group_diff_all("LIV", "BBS05", paste0("BBS0", c(1,2,4,7,8,9)), group2_name = "others")
```
*	Patients with BBS4 mutations are less likely to have liver anomalies compared to patients with other BBSome mutations.

```{r}
evaluators[["LIVER_BBS4_all"]] <-  evaluator_prob_group_diff_all("LIV", paste0("BBS0", c(1,2,5,7,8,9)), "BBS04", group1_name = "others")
```



```{r}
models_for_test <- all_models[c("gene_source","gene_source_filtered_lof","gene_imputed_age_sex")]

evaluation <- all_models %>% 
#evaluation <- models_for_test %>%
 map_df(function(def) {
    fit <- all_fits[[def$name]]
    data_for_prediction <- data_for_prediction_base_model(def, genes_to_show, phenotypes_to_show, age_transform = age_transform_from_age(data$age_numbers_groups_guessed))
    
    data_for_prediction_bbsome <- data_for_prediction %>% filter_for_BBSome()
    
    samples <- get_tidy_samples(fit, data_for_prediction)
    samples_diff <- get_samples_pairwise_diff(def, samples) %>% mutate(is_relevant = odds_ratio >= min_interesting_effect | odds_ratio <= 1 /min_interesting_effect)
    
    samples_diff_bbsome <- get_tidy_samples(fit, data_for_prediction_bbsome) %>% get_samples_pairwise_diff(def, .) %>% mutate(is_relevant = odds_ratio >= min_interesting_effect | odds_ratio <= 1 /min_interesting_effect)
    
    samples_prediction_bbsome = get_tidy_samples_prediction(fit, data_for_prediction_bbsome)
    
    samples_linear_bbsome <- get_tidy_samples(fit, data_for_prediction_bbsome, scale = "linear")
    
    lof_evaluation <- evaluator_lof(def, fit)
    
    evaluators %>% map(function(evaluator) {
      evaluator(samples_diff, samples_diff_bbsome, samples_prediction_bbsome, samples_linear_bbsome)
    }) %>% 
      do.call(cbind, .) %>% cbind(lof_evaluation, .) %>%
      mutate(model_name = def$name)
})

```


```{r load_freqentist}
#Freq. conclusions contain order
freq_conclusions <- readr::read_csv(here("private_data","BBS - Frequentist conclusions - p.csv"), col_types = cols(
  Conclusion = col_character(),
  `p-value-all` = col_double(),
  `p-value-cLOF` = col_double(),
  order = col_integer()
))


freq_conclusions <- freq_conclusions %>% 
  mutate(Conclusion = factor(Conclusion, levels = rev(Conclusion[order]))) %>%
  gather("model_name","p", `p-value-all`, `p-value-cLOF`) %>%
  mutate(model_nice = paste0("frequentist ", gsub("p-value-","", model_name)), Category = "Freq.")
```

```{r, fig.width = 8, fig.height = 4.5}
nice_model_name <- function(x)  {
  x %>% gsub("only","none", .) %>% gsub("gene_","",.) %>% gsub("filtered_","filtered ", .) %>%
    gsub("very_narrow","very narrow", .) %>% 
    gsub("lof_per_gene","cLOF (by gene)", .) %>% 
    gsub("lof","cLOF", .) %>% 
    gsub("cor"," corr.", .) %>% 
    gsub("imputed_","imputed ", .) %>%  gsub("_"," + ", .)
}

conclusion_names <- names(evaluation)[names(evaluation) != "model_name"]

if(!identical(sort(levels(freq_conclusions$Conclusion)), sort(conclusion_names))) {
  print(setdiff(levels(freq_conclusions$Conclusion), conclusion_names))
  print(setdiff(conclusion_names, levels(freq_conclusions$Conclusion)))
  stop("Conclusions mismatch")
}


evaluation_processed <- evaluation %>% 
  mutate(model_nice = factor(nice_model_name(model_name), levels = c(
   nice_model_name(names(all_models))
  ))) %>%
  gather("Conclusion","Probability", -model_name, -model_nice) %>%
  mutate(Conclusion = factor(Conclusion, levels = levels(freq_conclusions$Conclusion)),
         Category = case_when(model_name %in% c("gene_source","gene_source_filtered_lof") ~ "Main",
                              model_name %in% c("gene_only", "gene_lof", "gene_lof_per_gene", "gene_only_filtered_lof", "gene_source_very_narrow") ~ "Bad fits", 
                              TRUE ~ "Secondary") %>% factor(levels = c("Main","Secondary","Bad fits")))

build_conclusion_plot <- function(evaluation_processed, categories = TRUE) {
  if(categories) {
    facet <- facet_grid(.~Category, scales = "free_x", space = "free_x") 
  } else {
    facet <- NULL
  }

  evaluation_processed %>%
    ggplot(aes(x = model_nice, y = Conclusion, fill = Probability)) + geom_tile() + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust =0.5)) + 
      scale_fill_gradientn("Posterior prob.", colours = c("#1b7837","#f0f0f0", "#f0f0f0", "#762a83"),
                           values = c(0, 0.25,0.75,1),  limits = c(0,1)) +
      scale_x_discrete("Model components/modifications besides gene") +
      facet + base_theme
}

conclusions_plot <- build_conclusion_plot(evaluation_processed)
conclusions_plot

```

Heatmap of posterior probability of statements characterizing individual conclusions. All Bayesian models include gene as covariate but may also include additional covarietes: source, age, sex and loss of function (lof) - either as a global covariete or by gene. Since age and sex are not available for all data, we can either fit the model only to patients where those are reported (filtered) or impute missing data (imputed). Instead of using lof as a covariate, we can fit the model using only patients with lof mutations (filtered lof). For most models we include a correlation structure across phenotypes (e.g., that two phenotypes occur frequently together across all genes), but this structure may be absent (no corr.) or replaced with a correlation structure across genes (gene corr. - e.g., that two genes have similar pattern of effects across all phenotypes). We also tried modifying the width of prior distributions (wide, narrow, very narrow). See supplementary material for a detailed description of all models and the imputation procedure. Dark grey indicates that the question could not be evaluated for the given model (currently only asking for lof differences in models that exclude lof)



```{r}
build_freq_conclusions_plot <- function(freq_conclusions, categories = TRUE) {

  significance_pos <- 1 - (log10(0.05) / log10(min(freq_conclusions$p, na.rm = TRUE)))
  
  if(categories) {
    facet <- facet_grid(.~Category, scales = "free_x", space = "free_x") 
  } else {
    facet <- NULL
  }
  
  freq_conclusions %>% ggplot(aes(x = model_nice, y = Conclusion, fill = p)) + geom_tile() + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust =0.5)) + 
      scale_fill_gradientn("p-value", 
                           colours = c("#7f0000", "#fdd49e","#f0f0f0", "#f0f0f0"),
                           values = c(0, significance_pos - 0.01, significance_pos, 1), 
                           trans = "log10",
                           breaks = c(1e-8,1e-4,0.05), labels = c("1e-8","1e-4","0.05")) +
      scale_x_discrete("Freq. type") +
      expand_limits(fill = 1) +
      facet + base_theme
}

freq_conclusions_plot <-  build_freq_conclusions_plot(freq_conclusions)

freq_conclusions_plot
```


```{r, fig.width= 9, fig.height=4.5}
build_combined_conclusions_plot <- function(freq_conclusions_plot, conclusions_plot, rel_widths_plots, rel_widths_legend) {
  hide_guide_theme <- theme(legend.position = 'none')
  hide_y_axis_theme <- theme(axis.title.y = element_blank(), 
                             axis.text.y = element_blank(), 
                             axis.line.y = element_blank(),
                             axis.ticks.y = element_blank())
  
  plot_grid(
    plot_grid(freq_conclusions_plot + hide_guide_theme + theme(plot.margin = margin(7,0,7,7, "pt")), 
              conclusions_plot + hide_y_axis_theme + hide_guide_theme + theme(plot.margin = margin(7,7,7, 0, "pt")), 
              align = "h",rel_widths = rel_widths_plots),
    plot_grid(
      cowplot::get_legend(freq_conclusions_plot), 
      cowplot::get_legend(conclusions_plot),
      ncol = 1, nrow = 3),
    rel_widths = rel_widths_legend
  )
}

combined_conclusions_plot <- build_combined_conclusions_plot(freq_conclusions_plot, conclusions_plot, rel_widths_plots = c(1, 1.1), rel_widths_legend = c(4,1)) 
combined_conclusions_plot
```

```{r fig.width=7, fig.height=3.8}
hide_titles_theme <- theme(axis.title.x = element_blank())

combined_conclusions_main_plot <- build_combined_conclusions_plot(
  build_freq_conclusions_plot(freq_conclusions %>% filter(model_nice == "frequentist all") %>% mutate(model_nice = "frequentist")
                              , categories = FALSE) + hide_titles_theme,
  build_conclusion_plot(evaluation_processed %>% filter(model_nice == "source") %>% mutate(model_nice = "Bayesian"),
                        categories = FALSE) + hide_titles_theme,
  rel_widths_plots = c(9.6,1),
  rel_widths_legend = c(1,1)
)

combined_conclusions_main_plot
```


```{r}
combined_conclusions_plot %>% ggsave(here("tmp_pics","conclusions.pdf"), ., width = 9, height = 4.5)
combined_conclusions_main_plot %>% ggsave(here("tmp_pics","conclusions_main.pdf"), ., width = 7, height = 3.8)
```


```{r}
sessionInfo()
```

